Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-01-15 MANUSCRIPT.qmd
new figs and Neha edits

Directly measuring base flow and BFI presents unique challenges [@eckhardt2008]. The technique chosen to separate base flow has been shown to affect results, and the choice of base-flow separation method is subjective since ‘true’ BFI values are not known [@beck2013]. However, many methods have been developed to estimate these values. These methods include the use of tracers [@gonzales2009], graphical interpolation [@instituteofhydrology1980; @sloto1996], and digital filters [@arnold1995; @eckhardt2005; @lyne1979; @nathan1990]. These techniques have varying levels of applicability depending on the spatial scale, time span, and the scope of the study. Comparisons of various base-flow separation techniques have been made in [@eckhardt2005; @eckhardt2008; @nathan1990] , but this study does not explore the superiority of different methods.

Base flow was calculated using a single-parameter, recursive digital filter technique from [@nathan1990]. This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979]. @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study. The Lyne-Hollick filter has been used in [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018]. It takes the form of

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. This algorithm iteratively builds an ensemble of decision trees, where each tree corrects errors from previous trees to improve predictions [@chen2016xgboost]. Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].

@@

|              | Variable                  | Source       | Geoprocessing          |
|-----------------|---------------------|-----------------|------------------|
| Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
|              | Mean Temperature\*        | PRISM        | Basin average          |
|              | Mean Evapotranspiration\* | TerraClimate | Basin average          |
| Geospatial   | Elevation\*               | DEM          | Basin average          |
|              | Area                      | DEM          | Basin average          |
|              | Slope                     | DEM          | Basin average          |
|              | Aspect                    | DEM          | Percent areal coverage |
|              | Land Cover\*              | NLCD         | Percent areal coverage |
|              | Hydrologic Soil Group\*   | SSURGO       | Percent areal coverage |
|              | Geology                   | USGS         | Percent areal coverage |
|              | Karst                     | USGS         | Percent areal coverage |

@@

Feature extraction involves transforming the original dataset into a lower-dimensional feature space. However, this process generates new features that often lack the physical interpretability of the original variables. In contrast, feature selection identifies a subset of the original features, preserving their physical meaning while improving model readability and interpretability. In this study, supervised feature selection was employed to reduce the number of predictors, which enhanced learning performance, reduced computational costs, and mitigated overfitting.

@@

Statistical analyses were conducted on annual BFI and base-flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric]. This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations. This test is widely used in studies of this kind [@ficklin2016; @ayers2019; @woodhouse_udall22].

@@

Classifications presented in @tbl-trends were determined based on precipitation regime, physiographic region, climate, and slope. The dominant precipitation regime (monsoon vs. snowmelt) was identified by analyzing streamflow hydrographs for each station, focusing on peak flow periods during the monsoon season (July–September) and the snowmelt season (March–June). Physiographic region was assigned based on which region the streamgage is in. Climate classifications were defined as warm (above the median temperature of Arizona), cool (below the median temperature), wet (above the median precipitation), and dry (below the median precipitation). Slope was categorized as high (above the median slope) and low (below the median slope).

Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends). While decreasing trends dominate, both increasing and decreasing trends are observed within each classification. Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems may be more vulnerable to declining base flow. Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines. Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%). This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.

@@

The analysis shows that base flow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge. Coincidence with AET (actual evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods. In contrast, temperature trends show lower percentages of coincidence, often opposing base flow and BFI trends. Specifically, positive (negative) temperature trends are frequently associated with negative (positive) base-flow trends (60.98%) and BFI trends (52.68%). These results highlight the complex interplay between climatic variables and hydrological processes. While precipitation exerts the strongest influence on base flow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.

@@

: Coincident trends of climate variables (AET, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}

@@

Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted . The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values is acceptable (R^2^ = 0.764) indicating that the model performs well across the full dataset. The overall Root Mean Square Error (RMSE) is 0.129 and the overall percent bias (pbias) is -5.6. Model performance metrics across various classifications are summarized in @tbl-performance. These performance metrics demonstrate that the regional model performs consistently well across different spatial and climatic classifications. However, the negative pbias values across all classifications, along with the overall pbias, indicate a systematic underprediction of BFI by the model. Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those specific contexts.

@@

The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values. The most influential feature for predicting long-term BFI is basin elevation. While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013]. The importance of elevation aligns with findings from [@beck2013], highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration. Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022]. This relationship is further supported by hydroclimate features. Higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence. In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.

@@

While there has been an increase of stream gages in Arizona to comply with other imperatives, such as the Clean Water Act, these gages are not of sufficient quality for base-flow analyses. New, non-USGS stream gages are typically installed by flood control districts for flood flow monitoring, and do not accurately track base flow. New USGS stream gages have been added over the past 10 years to address instream flow rights and to improve the density of monitoring in the future.

>>>>

Directly measuring base flow and BFI presents unique challenges [@eckhardt2008]. The technique chosen to separate base flow has been shown to affect results, and the choice of base-flow separation method is subjective since ‘true’ BFI values are not known [@beck2013]. However, many methods have been developed to estimate these values. These methods include the use of tracers [@gonzales2009], graphical interpolation [@instituteofhydrology1980; @sloto1996], and digital filters [@arnold1995; @eckhardt2005; @lyne1979; @nathan1990]. These techniques have varying levels of applicability depending on the spatial scale, time span, and the scope of the study. Comparisons of various base-flow separation techniques have been made in previous studies [ e.g. @eckhardt2005; @eckhardt2008; @nathan1990]; this study does not explore the superiority of different methods.

Base flow was calculated using a single-parameter, recursive digital filter technique from @nathan1990. This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979]. @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study. The Lyne-Hollick filter has been used in multiple studies [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018], and it takes the form of

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors [@tbl-predictors]. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. This algorithm iteratively builds an ensemble of decision trees, where each tree corrects errors from previous trees to improve predictions [@chen2016xgboost]. Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].

@@

|              | Variable                  | Source       | Geoprocessing          |
|------------------|--------------------|------------------|------------------|
| Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
|              | Mean Temperature\*        | PRISM        | Basin average          |
|              | Mean Evapotranspiration\* | TerraClimate | Basin average          |
| Geospatial   | Elevation\*               | DEM          | Basin average          |
|              | Area                      | DEM          | Basin average          |
|              | Slope                     | DEM          | Basin average          |
|              | Aspect                    | DEM          | Percent areal coverage |
|              | Land Cover\*              | NLCD         | Percent areal coverage |
|              | Hydrologic Soil Group\*   | SSURGO       | Percent areal coverage |
|              | Geology                   | USGS         | Percent areal coverage |
|              | Karst                     | USGS         | Percent areal coverage |

@@

Feature extraction involves transforming the original dataset into a lower-dimensional feature space. However, this process generates new features that often lack the physical interpretability of the original variables. In contrast, feature selection identifies a subset of the original features, preserving their physical meaning while improving model readability and interpretability [@feat_selec2017]. In this study, supervised feature selection was employed to reduce the number of predictors, which enhanced learning performance, reduced computational costs, and mitigated overfitting.

@@

Statistical analyses were conducted on annual BFI and base-flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric]. This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations. This test is widely used in studies of this nature [@ficklin2016; @ayers2019; @woodhouse_udall22].

@@

Classifications presented in @tbl-trends were determined based on precipitation regime, physiographic region, climate, and slope. The dominant precipitation regime (monsoon vs. snowmelt) was identified by analyzing streamflow hydrographs for each station, focusing on peak flow periods during the monsoon season (July–September) and the snowmelt season (March–June). Physiographic region was assigned based on which region the streamgage is located. Climate classifications were defined as warm (above the long-term median temperature of Arizona), cool (below the long-term median temperature), wet (above the long-term median precipitation), and dry (below the long-term median precipitation). Slope was categorized as high (above the median slope) and low (below the median slope).

Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends). While decreasing trends dominate, both increasing and decreasing trends are observed within each classification. Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining base flow. Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines. Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%). This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.

@@

The analysis shows that base flow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge. Coincidence with ET~O~ (reference evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods. In contrast, temperature trends show lower percentages of coincidence, often opposing base flow and BFI trends. Specifically, positive (negative) temperature trends are frequently associated with negative (positive) base-flow trends (60.98%) and BFI trends (52.68%). These results highlight the complex interplay between climatic variables and hydrological processes. While precipitation exerts the strongest influence on base flow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.

@@

: Coincident trends of climate variables (ET~O~, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}

@@

Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted . The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values is acceptable (R^2^ = 0.764) indicating that the model performs well across the full dataset. The overall RMSE is 0.129 and the overall percent bias (pbias) is -5.6%. Model performance metrics across various classifications are summarized in @tbl-performance. These performance metrics demonstrate that the regional model performs consistently well across different spatial and climatic classifications. However, the negative pbias values across all classifications, along with the overall pbias, indicate a systematic underprediction of BFI by the model. Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those specific contexts.

@@

The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values. The most influential feature for predicting long-term BFI is basin elevation. While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013]. The importance of elevation aligns with findings from @beck2013, highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration. Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022]. This relationship is further supported by hydroclimate features where higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence. In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.

@@

While there has been an increase of stream gages in Arizona to comply with regulatory imperatives, such as the Clean Water Act, these gages are not of sufficient quality for base-flow analyses. New, non-USGS stream gages are typically installed by flood control districts for flood flow monitoring, and do not accurately track base flow. New USGS stream gages have been added over the past 10 years to address instream flow rights and to improve the density of monitoring in the future.
